Data Management for LSST Special Programs

### 2023-10-16 sec_intro.tex
update terms

Special Programs is a Rubin Data Management (RDM) term used to refer to survey 
components with specific sky regions within or beyond the 
Wide-Fast-Deep (WFD) main survey, and/or distinct cadences or 
strategies from the WFD main survey.

There are several requirements on the Rubin Data Management System (DMS) 
related to Special Programs, summarized in Section~\ref{sec:req}.
This document recommends implementation aspects for these requirements in order to 
maximize science from Special Programs (Section~\ref{sec:sci}).

Special Programs are driven by specific science goals that build on or 
add to the core science pillars of the LSST.
About 10--15\% of the total 10-year LSST time-span will be spent obtaining 
observations associated with Special Programs (and 85--90\% for WFD).

Special Programs includes non-WFD LSST components such as the Deep Drilling 
Fields (DDFs) and the mini-, micro-, and nano-surveys, some of which 
might use non-standard visits with different exposure times
(Section~\ref{ssec:intro_terms}).

\subsection{Additional Terms and Definitions}\label{ssec:intro_terms}

\textbf{Wide-Fast-Deep (WFD) Main Survey -- }
The core science program of the LSST, designed to achieve the science 
goals defined by the Science Requirements Document (SRD; \citeds{LPM-17}).
It is a contiguous area that covers at least $\sim$18000 deg$^2$ with 
$\gtrsim$800 visits per field, expected to be accomplished with 85--90\% 
of the observing time.

As of the Phase 2 SCOC recommendations in \citeds{PSTN-055}, the WFD area 
is considered to include the low dust extinction extragalactic regions; 
the higher extinction, high stellar density Galactic Plane (GP) and Bulge; the 
Virgo Cluster; the South Celestial Pole (SCP); and the North Ecliptic Spur 
(NES).

In the past the Galactic, SCP, and NES regions have been referred to as 
``mini-surveys" (see below) but this does not make them "Special Programs". 
From a survey-scheduling perspective, they are now considered part of the 
WFD survey (whether or not the scheduler software treats them as such).

From a science and data-processing perspective, these components of the 
WFD do not require special processing or separate data products 
(i.e., do not need to be treated by RDM as ``Special Programs").
This is because they will use standard (or alternative) visits and can
be incorporated into WFD Prompt processing (Section~\ref{ssec:proc_wfd}), 
and because their distinct areas do not overlap with the other WFD components 
and so do not require, e.g., separate deep coadds.

\textbf{Deep Drilling Field (DDF) -- }
A single pointing for which many (e.g., a hundred) visits are obtained 
(usually sequentially) during a single night, and repeated every few 
nights.

@@

From a science and data-processing perspective, the DDFs do require 
special processing and separate data products (Section~\ref{ssec:proc_rdm}).

\textbf{Mini-, Micro- and Nano-Surveys -- }
Specific sky areas covered by a few hundred, a hundred, or tens of visits 
(respectively; as in Phase 2 SCOC recommendations, \citeds{PSTN-055}).
For a list of the current mini-surveys under consideration, see 
\citeds{PSTN-055}.
This document refers to them all together as mini-surveys.

These sky areas are all (so far) either within or adjacent to the WFD area.
In some cases these visits are non-standard (e.g., short exposures).

From a data-processing and science perspective, most mini-surveys are 
likely to require special processing or separate data products, and should 
be considered on a case-by-base basis.

\textbf{Visit Types -- }
A visit is an observation of a single pointing at a given time, of which 
there are three types:

The Phase 2 SCOC recommendations for the WFD (\citeds{PSTN-055}) are to 
use alternative standard visits in the $u$-band and standard visits in all 
other bands; variable exposure times were not recommended for the WFD. 
Non-standard visits with shorter or longer exposure times are being 
considered for some mini-surveys.

>>>>

The three memoranda for Rubin Data Management regarding Special Programs are:
\begin{enumerate}
\item All visits should have both a region and a program label.
These labels should be propagated to all derived data products
(e.g., processed images, catalog data, and alerts; Section~\ref{ssec:sci_labels}).
\item All visits that \emph{can} be processed by the Prompt pipelines and generate 
alerts \emph{should} be, in support of time domain and Solar System science goals (Section~\ref{ssec:proc_wfd}).
\item Special Processing to create ``unique and separate" data products for Special Programs
should be done when it is possible and necessary, as defined in Section~\ref{ssec:proc_rdm}. 
\end{enumerate}

\subsection{Terms and Definitions}\label{ssec:intro_terms}

\subsubsection{Visit Types}

A visit is an observation of a single pointing at a given time, of which 
there are three types:

Non-standard visits with shorter or longer exposure times are being 
considered for some Special Programs.

\subsubsection{Wide Fast Deep (WFD) Main Survey}

The WFD program is the core science program of the LSST, designed to achieve the science 
goals defined by the Science Requirements Document (SRD; \citeds{LPM-17}).

The WFD is defined from a scheduling perspective as a certain cadence and filter balance 
(i.e., an observing strategy; a program) 
that covers the low-dust region (extragalactic areas) and several special regions:
the Galactic bulge and plane; the South Celestial Pole (SCP); the North Ecliptic Spur (NES);
and the Virgo Cluster \citedsp{PSTN-055}.

The WFD's regions will compose a contiguous sky footprint that covers at least $\sim$18000 deg$^2$ with 
$\gtrsim$800 visits per field, expected to be accomplished with 85--90\% 
of the observing time.

The WFD is expected to use alternative standard visits in the $u$-band and 
standard visits in all other bands \citedsp{PSTN-055}.
Rolling cadence is expected to be implemented in the low-dust region only.

\subsubsection{Regular (non-special) processing}

The image processing and data products described in the Data Products Definitions Document
\citedsp{LSE-163} that are designed for, and will be applied to, the WFD program's observations.

In some cases, regular processing is also appropriate for Special Programs.

\subsubsection{Special Programs}

This is a Rubin Data Management (RDM) term used to refer to sky regions within
or beyond the WFD footprint which have a distinct observing strategy
(e.g., visit type, cadence, filter distribution).

Special Programs are typically driven by specific science goals that build on or 
add to the core science pillars of the LSST.
About 10--15\% of the total 10-year LSST time-span will be spent obtaining 
observations associated with Special Programs.
Special Programs includes non-WFD LSST components such as the Deep Drilling 
Fields (DDFs) and the mini-, micro-, and nano-surveys \citedsp{PSTN-055}.

Some science goals for Special Programs can be met with regular (non-special) processing,
but some will require Special Processing by Rubin Data Management or user-generated processing.

\subsubsection{Special Processing}

The data processing pipelines, built from the existing LSST Science Pipelines, which are
applied by Rubin Data Management to observations from Special Programs.
Special Processing creates data products that are unique and separate from those produced
by regular processing for the WFD program.

Special Processing is likely to use special inputs, configurations, 
or timescales, as appropriate for the Special Programs' data and science goals.
However, the development and application of specialized \emph{algorithms} or new software
is beyond the scope of Special Processing by Rubin Data Management.

\subsubsection{User-Generated Processing}

Any processing of Rubin data done by users in order to reach specific science goals, including
processing for Special Programs data, is referred to as user-generated processing.

Guidelines for user-generated processing that can be run in the 
Rubin Science Platform, and for user-generated data products that can
be federated with the Rubin-product data products (i.e., joinable tables),
is forthcoming.

\subsubsection{Deep Drilling Field (DDF)}

A single pointing for which many (e.g., a hundred) visits are obtained 
(usually sequentially) during a single night, and repeated every few 
nights.

@@

The DDFs will require Special Processing and unique and separate data products (Section~\ref{ssec:proc_rdm}).

\subsubsection{Mini-, Micro- and Nano-Surveys}

Specific sky areas covered by a few hundred, a hundred, or tens of visits (respectively).
This document will refer to them collectively as mini-surveys.

The sky areas of mini-surveys can be within, adjacent to, or detached from the WFD footprint.
Mini-surveys can have non-standard visits.

For a list of the mini-surveys under consideration,
see \citeds{PSTN-055}).

Most mini-surveys are likely to require Special Processing and unique and separate data products,
and should be considered on a case-by-base basis.
Some mini-surveys will likely require user-generated processing.
